---
title: "Cyclistic Member vs Casual Rider Useage Pattern"
author: "Muhammad Ahmed"
date: "3rd of October 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project was undertaken as part of the Google Professional Certification, Introduction to Data Analytics. Let's dive in!

## Introduction
Initially I had thought to clean data in SQL and then use Tableau for graphics,
however, due to the limits in BigQuery and Tableau Public, I decided to use R

## Background
I am a junior data analyst working on the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, my team wants to understand how casual riders and annual members use Cyclistic bikes differently. From theseinsights, my team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations

## ASK
In this regards, the marketing department is tasked with 3 questions to answer
*How do annual members and casual riders use Cyclistic bikes differently?
*Why would casual riders buy Cyclistic annual memberships?
*How can Cyclistic use digital media to influence casual riders to become members?

I am assigned a task my the marketing department director(Lily Moreno) : How do annual members and casual riders use Cyclistic bikes differently?

## Prepare DATA

### Loading necessary libraries
```{r loads necessary libraries}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
library(ggplot2)
```

### Loading Files

Once libraries have been loaded, it is time to upload the CSV files with data in it. Each month has been assigned a number(M). The data is from 2024-06 onwards. Once uploaded, all data frames are combined into a single data frame with the name of bike_rides. We also check the dimensions of this new data frame

```{r loading files}
###Working directory was set from GUI, this was to confirm
getwd()

###Read CSV Function for importing files, each month assigned as M(number)
###Latest data first, 202406 and onwards had time in different format
M1 <- read.csv('202405-divvy-tripdata.csv', header=TRUE)
M2 <- read.csv('202404-divvy-tripdata.csv', header=TRUE)
M3 <- read.csv('202403-divvy-tripdata.csv', header=TRUE)
M4 <- read.csv('202402-divvy-tripdata.csv', header=TRUE)
M5 <- read.csv('202401-divvy-tripdata.csv', header=TRUE)
M6 <- read.csv('202312-divvy-tripdata.csv', header=TRUE)
M7 <- read.csv('202311-divvy-tripdata.csv', header=TRUE)
M8 <- read.csv('202310-divvy-tripdata.csv', header=TRUE)
M9 <- read.csv('202309-divvy-tripdata.csv', header=TRUE)
M10 <- read.csv('202308-divvy-tripdata.csv', header=TRUE)
M11 <- read.csv('202307-divvy-tripdata.csv', header=TRUE)
M12 <- read.csv('202306-divvy-tripdata.csv', header=TRUE)

####Now this might not be an efficient method but for now, this is how I did it

###Combined all data frames into one for analysis
bike_rides <- rbind(M1,M2,M3,M4,M5,M6,M7,M8,M9,M10,M11,M12)

###Checking the dimensions of this whole thing
dim(bike_rides)
```

### Cleaning Data

The following code is run to clean data. First, empty columns and rows are removed.

The start and end times are converted into the ymd_hms format (for calculations). The difftime function is used to calculate the time difference. To ensure that data is upto standards, we remove non finite values as well as values below 1 minute (negative and start stops) as well as longer than a day (user forgot to turn off bike etc.)

We also removed the data where start and end stations are missing.

We also extract the months and days from the data (as we will be using this further) and assign them to a new column of month and day respectively. In the end, all the processed and clean data is shifted into a new data frame bike_rides_clean.

```{r cleaning data}
###This part removes empty columns and rows
bike_rides <- janitor::remove_empty(bike_rides,which=c("cols"))
bike_rides <- janitor::remove_empty(bike_rides,which=c("rows"))

###Converting Data/Time from chr to Date/Time format and assigning new columns
bike_rides$start_time <- lubridate::ymd_hms(bike_rides$started_at)
bike_rides$end_time <- lubridate::ymd_hms(bike_rides$ended_at)

###Calculating the difference in time
bike_rides$ride_time <- difftime(bike_rides$end_time,bike_rides$start_time,units="mins")

###Remove rows where ride_time has non-finite values (NA, NaN, Inf)
bike_rides_clean <- bike_rides %>%
  filter(is.finite(ride_time))

###Ensure the ride_time column is numeric (difftime can sometimes cause issues with plotting)
bike_rides$ride_time <- as.numeric(bike_rides$ride_time)

bike_rides <- bike_rides %>%
  filter(!is.na(start_station_name) & start_station_name != "",
         !is.na(end_station_name) & end_station_name != "")

###Removing Bike Rides less than a minute and more than a day
bike_rides_clean <- bike_rides %>%
  filter(ride_time >= 1 & ride_time <= 1440)

###Check the summary to understand the range of ride_time
summary(bike_rides_clean$ride_time)

###Month and Year extracted from start time for analysis
bike_rides_clean$ride_month <- format(bike_rides_clean$start_time, "%Y-%m")
###Extract the day of the week from the start time
bike_rides_clean$day_of_week <- wday(bike_rides_clean$start_time, label = TRUE, abbr = FALSE) 

###Checking Dimensions again to see if any empty rows or columns were removed
dim(bike_rides_clean)

###Count the number of rides per month for each user type
rides_per_month <- bike_rides_clean %>%
  group_by(ride_month, member_casual) %>%
  summarise(total_rides = n()) %>%
  ungroup()

```


## Analyzing the DATA

### Daily Rides Breakdown

This chunk of code is used to count the number of rides per day of the week for each user type
```{r}
rides_per_day <- bike_rides_clean %>%
  group_by(day_of_week, member_casual) %>%
  summarise(total_rides = n()) %>%
  ungroup()
```
View the result
```{r}
print(rides_per_day)
```
This chunk of code gives the plot for the the number of rides per day of the week for each user type
```{r Daily Rides Breakdown, echo=FALSE}
color_palette <- c("member" = "#CE5757", "casual" = "#99CCFF")  # Pastel red and blue
ggplot(rides_per_day, aes(x = day_of_week, y = total_rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  ####Dodged bar plot for better comparison
  labs(title = "Number of Rides per Day of the Week by User Type",
       x = "Day of the Week",
       y = "Total Rides") +
  scale_fill_manual(values = color_palette) + 
  theme_minimal()
```

### Monthly Breakdown - Riders Type

```{r Monthly Breakdown - Riders Type}
###View result
print(rides_per_month)

###Define pastel colors for types of users
color_palettes <- c("member" = "#CE5760", "casual" = "#99CCFF")  # Pastel red and blue

```
The results are as follows:
```{r Monthly Breakdown by Riders Type, echo=FALSE}
###Plot the number of rides per month for each user type
ggplot(rides_per_month, aes(x = ride_month, y = total_rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Number of Rides per Month by User Type",
       x = "Month",
       y = "Total Rides") +
  scale_fill_manual(values = color_palettes) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Time spent riding bikes by user type
```{r Ride Time, echo=FALSE}
ggplot(bike_rides_clean, aes(x = ride_time, fill = member_casual)) +
  geom_histogram(binwidth = 5, position = "stack", color = "black") +
  labs(title = "Distribution of Ride Times by Membership Type",
       x = "Ride Time (minutes)",
       y = "Frequency") +
  xlim(0, quantile(bike_rides$ride_time, 0.97)) +  ####Set limits to exclude outliers beyond 97th percentile
  scale_fill_manual(values = color_palette) + 
  theme_minimal()
```

### Bike Ridership at different hours of the day

Extract the hour of the start time for analysis
```{r Hour of Day}
bike_rides_clean$start_hour <- hour(bike_rides_clean$start_time)
```
Data Plot
```{r Hour of day, echo=FALSE}
###Plot the stacked histogram by start time (hour)
ggplot(bike_rides_clean, aes(x = start_hour, fill = member_casual)) +
  geom_histogram(binwidth = 1, position = "stack", color = "black") +  
  labs(title = "Ride Start Times by Membership Type",
       x = "Hour of the Day",
       y = "Frequency") +
  scale_y_continuous(labels = scales::label_number()) +  
  scale_x_continuous(breaks = 0:23, limits = c(0, 24)) + 
  scale_fill_manual(values = color_palette) + 
  theme_minimal()
```

### Most Frequented Stations by user type
The following code is used to differentiate the most used start and stop stations, filtered by month and user type
```{r}

###Group by month, user type, and start station, count rides
most_used_start_stations <- bike_rides_clean %>%
  group_by(ride_month, member_casual, start_station_name) %>%
  summarise(total_rides = n()) %>%
  arrange(ride_month, member_casual, desc(total_rides)) %>%
  group_by(ride_month, member_casual) %>%
  slice_max(total_rides, n = 5)  ####Top 5 stations

###Group by month, user type, and end station, count rides
most_used_end_stations <- bike_rides_clean %>%
  group_by(ride_month, member_casual, end_station_name) %>%
  summarise(total_rides = n()) %>%
  arrange(ride_month, member_casual, desc(total_rides)) %>%
  group_by(ride_month, member_casual) %>%
  slice_max(total_rides, n = 5)  ####Top 5 stations

# Define 11 pastel blue colors
pastel_blue_colors <- c(
  "#A3C1DA", "#A0C4E1", "#A8D5E2", "#B7E1E5", "#C6E4E7",
  "#B2D9E4", "#9EC9E5", "#8FB7D3", "#77A6C8", "#3F87A6",
  "#0A9AC3", "#0A8AD3", "#016CC3", "#024CC3", "#011CC1"
)

### Start station plot for casual riders with horizontal facet labels
ggplot(most_used_start_stations %>% filter(member_casual == "casual"), 
       aes(x = "", y = total_rides, fill = start_station_name)) +
  geom_bar(stat = "identity", width = 1, color = "black") + 
  facet_wrap(~ride_month, ncol = 1, strip.position = "right") + 
  coord_flip() +
  labs(title = "Top 5 Start Stations by Casual Riders Each Month",
       y = "Date (Year-Month)",
       x = NULL) +
  scale_fill_manual(values = pastel_blue_colors) + 
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0), 
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right"
  )

### End station plot for casual riders with horizontal facet labels
ggplot(most_used_end_stations %>% filter(member_casual == "casual"), 
       aes(x = "", y = total_rides, fill = end_station_name)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  facet_wrap(~ride_month, ncol = 1, strip.position = "right") +
  coord_flip() +
  labs(title = "Top 5 End Stations by Casual Riders Each Month",
       y = "Date (Year-Month)",
       x = NULL) +
  scale_fill_manual(values = pastel_blue_colors) +  
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0), 
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right"
  )

### Start station plot for members with horizontal facet labels
ggplot(most_used_start_stations %>% filter(member_casual == "member"), 
       aes(x = "", y = total_rides, fill = start_station_name)) +
  geom_bar(stat = "identity", width = 1, color = "black") + 
  facet_wrap(~ride_month, ncol = 1, strip.position = "right") + 
  coord_flip() +
  labs(title = "Top 5 Start Stations by Members Each Month",
       y = "Date (Year-Month)",
       x = NULL) +
  scale_fill_manual(values = pastel_blue_colors) + 
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0), 
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right"
  )

### End station plot for members with horizontal facet labels
ggplot(most_used_end_stations %>% filter(member_casual == "member"), 
       aes(x = "", y = total_rides, fill = end_station_name)) +
  geom_bar(stat = "identity", width = 1, color = "black") +  
  facet_wrap(~ride_month, ncol = 1, strip.position = "right") + 
  coord_flip() +
  labs(title = "Top 5 End Stations by Members Each Month",
       y = "Date (Year-Month)",
       x = NULL) +
  scale_fill_manual(values = pastel_blue_colors) + 
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right"
  )


```

#### Stations as percentage of Total rides in given month

```{r Monthly Stations Percentage}
###First, calculate the total rides per month for casual and member riders
total_rides_per_month <- bike_rides_clean %>%
  group_by(ride_month, member_casual) %>%
  summarize(total_rides_month = n())

###Calculate percentage for top 5 start stations
most_used_start_stations_percent <- most_used_start_stations %>%
  left_join(total_rides_per_month, by = c("ride_month", "member_casual")) %>%
  mutate(percent_rides = (total_rides / total_rides_month) * 100)

###Calculate percentage for top 5 end stations
most_used_end_stations_percent <- most_used_end_stations %>%
  left_join(total_rides_per_month, by = c("ride_month", "member_casual")) %>%
  mutate(percent_rides = (total_rides / total_rides_month) * 100)
```
Plotting the results
```{r Monthly Station Percentage, echo=FALSE}
# Define 14 pastel blue colors
pastel_blue_colors <- c(
  "#A3C1DA", "#A0C4E1", "#A8D5E2", "#B7E1E5", "#C6E4E7",
  "#B2D9E4", "#9EC9E5", "#8FB7D3", "#77A6C8", "#3F87A6",
  "#0A8CC3", "#0091B9", "#0072A8", "#005B88"
)


### Plot the start stations for casual riders with percentages and gaps
ggplot(most_used_start_stations_percent %>% filter(member_casual == "casual"), 
       aes(x = percent_rides, y = ride_month, fill = start_station_name)) +
  geom_bar(stat = "identity", width = 0.7, position = position_stack(), color = "black") +  
  labs(title = "Percentage of Total Rides from Top 5 Start Stations (Casual Riders)",
       x = "Percentage of Total Rides",
       y = "Date (Year-Month)") +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  scale_fill_manual(values = pastel_blue_colors) +  # Use the defined pastel blue colors
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "right"
  )

### Plot the end stations for casual riders with percentages and gaps
ggplot(most_used_end_stations_percent %>% filter(member_casual == "casual"), 
       aes(x = percent_rides, y = ride_month, fill = end_station_name)) +
  geom_bar(stat = "identity", width = 0.7, position = position_stack(), color = "black") +  
  labs(title = "Percentage of Total Rides from Top 5 End Stations (Casual Riders)",
       x = "Percentage of Total Rides",
       y = "Date (Year-Month)") +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) + 
  scale_fill_manual(values = pastel_blue_colors) +  # Use the defined pastel blue colors
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "right"
  )

### Plot the start stations for members with percentages and gaps
ggplot(most_used_start_stations_percent %>% filter(member_casual == "member"), 
       aes(x = percent_rides, y = ride_month, fill = start_station_name)) +
  geom_bar(stat = "identity", width = 0.7, position = position_stack(), color = "black") + 
  labs(title = "Percentage of Total Rides from Top 5 Start Stations (Members)",
       x = "Percentage of Total Rides",
       y = "Date (Year-Month)") +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) + 
  scale_fill_manual(values = pastel_blue_colors) +  # Use the defined pastel blue colors
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "right"
  )

### Plot the end stations for members with percentages and gaps
ggplot(most_used_end_stations_percent %>% filter(member_casual == "member"), 
       aes(x = percent_rides, y = ride_month, fill = end_station_name)) +
  geom_bar(stat = "identity", width = 0.7, position = position_stack(), color = "black") + 
  labs(title = "Percentage of Total Rides from Top 5 End Stations (Members)",
       x = "Percentage of Total Rides",
       y = "Date (Year-Month)") +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +  
  scale_fill_manual(values = pastel_blue_colors) +  # Use the defined pastel blue colors
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "right"
  )


##Type of ride used by each member type ----------
###Calculate the count of each rideable_type by member_casual
bike_usage_by_type <- bike_rides_clean %>%
  group_by(member_casual, rideable_type) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  group_by(member_casual) %>%
  mutate(percent = count / sum(count) * 100)

###Define pastel color palette for rideable types
rideable_color_palette <- c("electric_bike" = "#A2D1E6",  
                            "docked_bike" = "#DDF0F7",    
                            "classic_bike" = "#769FCA") 

###Create a function to plot pie charts
plot_pie_chart <- function(data, user_type) {
  ggplot(data, aes(x = "", y = percent, fill = rideable_type)) +
    geom_bar(stat = "identity", width = 1, color = "black") +
    coord_polar("y") +
    labs(title = paste("Percentage of Rideable Types -", user_type)) +
    scale_fill_manual(values = rideable_color_palette) +  # Manual color scale for rideable types
    theme_void() +
    theme(legend.position = "right")  # Adjust legend position
}

###Separate data for members and casual riders
members_data <- bike_usage_by_type %>% filter(member_casual == "member")
casual_data <- bike_usage_by_type %>% filter(member_casual == "casual")

###Plot pie chart for members
plot_pie_chart(members_data, "Members")

###Plot pie chart for casual riders
plot_pie_chart(casual_data, "Casual Riders")



```

### Bike type used by each user type
The following code is used to find the useage of each bike type by the two types of users (member and casual)
```{r Bike Type by user}
###Calculate the count of each rideable_type by member_casual
bike_usage_by_type <- bike_rides_clean %>%
  group_by(member_casual, rideable_type) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  group_by(member_casual) %>%
  mutate(percent = count / sum(count) * 100)

###Define pastel color palette for rideable types
rideable_color_palette <- c("electric_bike" = "#A2D1E6",  
                            "docked_bike" = "#DDF0F7",    
                            "classic_bike" = "#769FCA") 

###Create a function to plot pie charts
plot_pie_chart <- function(data, user_type) {
  ggplot(data, aes(x = "", y = percent, fill = rideable_type)) +
    geom_bar(stat = "identity", width = 1, color = "black") +
    coord_polar("y") +
    labs(title = paste("Percentage of Rideable Types -", user_type)) +
    scale_fill_manual(values = rideable_color_palette) +  # Manual color scale for rideable types
    theme_void() +
    theme(legend.position = "right")  # Adjust legend position
}

###Separate data for members and casual riders
members_data <- bike_usage_by_type %>% filter(member_casual == "member")
casual_data <- bike_usage_by_type %>% filter(member_casual == "casual")

```
Plotting the results
```{r Bike type by user, echo=FALSE}
###Plot pie chart for members
plot_pie_chart(members_data, "Members")

###Plot pie chart for casual riders
plot_pie_chart(casual_data, "Casual Riders")

```
